Large squeezing in multicomponent trapped Bose-Einstein condensates

### 2012-12-17 large_number_squeezing_with_atoms.tex
Starting work on the Bog section

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass{iopart}
%\newcommand{\gguide}{{\it Preparing graphics for IOP journals}}
%Uncomment next line if AMS fonts required
%\usepackage{iopams}
\usepackage{graphicx}
\usepackage{ulem}
\usepackage{amssymb}
\usepackage{caption}
\captionsetup{compatibility=false}
\usepackage{subcaption}

@@

\section{Squeezing in 2D and 3D}
The simulations in the previous two sections were one dimensional, and show how having spatially constant density and using a $\pi$-pulse will improve the resultant squeezing.  While these conclusions remain valid for simulations in two or three dimensions, adding each additional dimension results in a degradation of the squeezing.  This is true even when the potential (and the resulting atomic density) is constant across the bulk of the trap.  Furthermore, the expected squeezing goes down as the volume of the trap increases.  This can be seen in Fig.~\ref{figDimensionalSqueezingEffects}, where we show the squeezing generated in 1D, 2D and 3D BECs for two different parameter regimes. In both cases we take the BEC to be trapped in a box potential, which in the Thomas-Fermi limit leads to a constant density profile across the BEC. In Figure~\ref{figDimensionalSqueezingEffects}a we consider the case where $U_{aa}=U_{bb}\neq 0$, $U_{ab}=0$ $U_{aa}\neq 0$, while in Figure~\ref{figDimensionalSqueezingEffects}b we consider the case $U_{aa}\neq 0$, $U_{ab}=U_{bb}=0$. In both cases it is clear that number squeezing is best in lower dimensions, and becomes degraded or non-existent as we move to a three dimensional system; the effect is worse if more than one nonlinearity is non-zero.

@@

Excitations of a BEC are described by the Bogoliubov modes...

@@

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass{iopart}
%\newcommand{\gguide}{{\it Preparing graphics for IOP journals}}
%Uncomment next line if AMS fonts required
%\usepackage{iopams}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{ulem}
\usepackage{amssymb}
\usepackage{caption}
\captionsetup{compatibility=false}
\usepackage{subcaption}

\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\sinc}{sinc}
\newcommand{\abs}[1]{\ensuremath{\left|#1\right|}}
\newcommand{\expect}[1]{\ensuremath{\left<#1\right>}}
\newcommand{\ket}[1]{\ensuremath{\left|#1\right>}}

@@

\section{Squeezing in 2D and 3D}
The simulations in the previous two sections were one dimensional, and show how having spatially constant density and using a $\pi$-pulse will improve the resultant squeezing.  While these conclusions remain valid for simulations in two or three dimensions, adding each additional dimension results in a degradation of the squeezing.  This is true even when the potential (and the resulting atomic density) is constant across the bulk of the trap.  Furthermore, the expected squeezing goes down as the volume of the trap increases.  This can be seen in Fig.~\ref{figDimensionalSqueezingEffects}, where we show the squeezing generated in 1D, 2D and 3D BECs for two different parameter regimes. In both cases we take the BEC to be trapped in a box potential, which in the Thomas-Fermi limit leads to a constant density profile across the BEC. In Figure~\ref{figDimensionalSqueezingEffects}a we consider the case where $U_{aa}=U_{bb}\neq 0$, $U_{ab}=0$, while in Figure~\ref{figDimensionalSqueezingEffects}b we consider the case $U_{aa}\neq 0$, $U_{ab}=U_{bb}=0$. In both cases it is clear that number squeezing is best in lower dimensions, and becomes degraded or non-existent as we move to a three dimensional system; the effect is worse if more than one nonlinearity is non-zero.

@@

We can understand the origin and scaling of this degradation of squeezing by considering the response of the condensate to small fluctuations about the mean field.  We do this using Bogoliubov theory \cite{PethickSmith}, which considers the first order quantum-mechanical fluctuations about the mean field.  To simplify the analysis we consider the case in which $U_{aa} \neq 0$, $U_{bb}=U_{ab} = 0$ and the condensate has a spatially-constant density in a box of side lengths $L_x$, $L_y$ and $L_z$.  In this limit the squeezing of the $a$ mode at $t_3$ is
\begin{align}
  \left.\frac{\Var[N_a](t_3)}{N_a(t_3)}\right|_\text{Bog} &= \frac{\Var[N_a](t_3) + \sum_\mathbf{k} \frac{1}{2} n_a(\mathbf{k}, t_3) \left\{3 + \cos\left[\Omega(t_3-t_2)\right] + 4 n_a(\mathbf{k}, t_3)\right\}}{N_a(t_3) + \sum_\mathbf{k} n_a(\mathbf{k}, t_3)},\\
  n_a(\mathbf{k}, t_3) &= \left\{n_a \lambda_{aa} \cos\left[\Omega\left(t_3-t_2\right)\right] \sinc\left(\omega_\mathbf{k} \tau_\text{hold}\right) \right\}^2,\\
  \omega_\mathbf{k} &= \sqrt{\omega^0_\mathbf{k}(\omega^0_\mathbf{k} + 2 \lambda_{aa} n_a)}, \\
  \omega^0_\mathbf{k} &= \frac{\hbar \mathbf{k}^2}{2 M},
\end{align}
where $n_a(\mathbf{k}, t) = \expect{\delta \hat{a}^\dagger(t) \delta\hat{a}(t)}$ is the occupation of momentum mode $\mathbf{k}$, $\Var[N_a](t)$ and $N_a(t)$ are the number variance and mean number of the $\ket{a}$ mode from the single-mode derivation, the sum over $\mathbf{k}$ is a discrete sum over the available momentum modes $\mathbf{k} = n_x/2\pi L_x \hat{\mathbf{x}} + n_y/2\pi L_y\hat{\mathbf{y}} + n_z/2\pi L_z\hat{\mathbf{z}}$, $\omega_\mathbf{k}$ is the frequency of the Bogoliubov mode $\mathbf{k}$ and $\omega^0_\mathbf{k}$ is the frequency of a free particle.

@@

\section{Derivation of Bogoliubov squeezing formula}
\label{appendixBogDerivation}

Here we put some details of the Bog analysis.

\begin{align}
  S &= \frac{\Var(n_0) + \sum_k \frac{1}{2} n_k (3 + \cos 2\theta + 4 n_k)}{n_0 + \sum_k n_k} \\
  n_k &= \left[\lambda \cos\theta \sinc(\omega_k \tau)\right]^2 \\
  \hat{a}(t_3) &= \left[\hat{a}(0) - i \chi \alpha \tau_h\left(\alpha^* \hat{a}(0) + \alpha \hat{a}^\dagger(0) - 2\abs{\alpha}^2\right) \right]\cos(\theta) - i\hat{b}(0) \sin\theta e^{i\phi} \\
  n_0 &= \left(\alpha ^2+\alpha ^4 \lambda ^2\right) \cos^2\theta + \left(\beta^2 \sin\theta  + 2\alpha\beta\cos\theta\sin\phi\right)\sin\theta\\
  \begin{split}
  \Var(n_0) &= \frac{1}{2} \left(\alpha ^2+\beta ^2\right)+\frac{1}{8}\alpha ^4 \left(7+ 2\beta ^2\right) \lambda ^2+ \frac{3}{4} \alpha ^8 \lambda ^4 \\
  &\phantom{\mathrel{=}}+\frac{1}{2}\left( \alpha ^2-\beta ^2+2 \alpha ^4 \lambda ^2+2 \alpha ^8 \lambda ^4\right) \text{Cos}[2 \theta ] \\
  &\phantom{\mathrel{=}}+\frac{1}{8}\left(\alpha ^4 \lambda ^2 \left(1-2 \beta ^2+2 \alpha ^4 \lambda ^2\right) \text{Cos}[4 \theta ] \right) \\
  &\phantom{\mathrel{=}}+\frac{1}{8}\left(4 \alpha  \beta  \left(8 \alpha ^2 \lambda  \text{Cos}[\theta ]^3 \text{Cos}[\phi ] \text{Sin}[\theta ]+\text{Sin}[2 \theta ] \left(2 \text{Sin}[\phi ]+\alpha  \beta  \lambda  \text{Sin}[2 \theta ] \left(\alpha ^2 \lambda  \text{Cos}[2 \phi ]+\text{Sin}[2 \phi ]\right)\right)\right)\right)
  \end{split}
\end{align}

@@

\bibitem{PethickSmith} C.J.~Pethick and H.~Smith, \emph{Bose-Einstein Condensation in Dilute Gases} (Cambridge University Press, Cambridge, 2002).
